---
title: "Gapminder_Income_EDA"
author: "Logeshwar Venkatesan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    fig_caption: true
    theme: flatly
---
      
# Target of this analysis
    
In this analysis, we will look at the variable income and understand how it affects three other important variables viz., population, life expectancy and geography.
  
## Data exploration
  
Load csv and read in data
  
```{r,echo=TRUE,warning=TRUE}  
    library(readr)  
    gapminder<-read_csv('gapminder.csv', col_types = 'cnnnnc')
```
  
## Check packaging

Check number of rows and columns in data set
  
```{r,echo=TRUE,warning=TRUE}
    nrow(gapminder)
    ncol(gapminder)
```
  
## Run str()
  
Analyze the internal structure of the R object
  
```{r,echo=TRUE,warning=TRUE}
    str(gapminder)
```  
Look at the top and bottom of the data to check formatting errors
  
```{r,echo=TRUE,warning=TRUE}
    head(gapminder)
    tail(gapminder)
```
## Check missing data
  
### Count of missing values

```{r,echo=TRUE,warning=TRUE}
    sapply(gapminder, function(x) sum(is.na (x)))
```
##  Data aggregation
```{r,echo=TRUE,warning=TRUE}
    library(dplyr)
    gapminder %>%
    group_by(Country) %>%
    summarise(life = median(life))
```
  
## Plotting

### Total Dataset

In this section, we will plot a graph to understand the overall dataset and in the later sections, we will look at the questions we formulated.
  
```{r,echo=TRUE,warning=TRUE}
    library(dplyr)
    table(gapminder$income) %>% barplot(col = "wheat")
    table(gapminder$population) %>% barplot(col = "wheat")
    table(gapminder$life) %>% barplot(col = "wheat")
```

### Finite Dataset
  
**The analysis conducted is only after the year 2000 and hence we'll filter the data accordingly**
  
Filter data from main dataset to include data for only after year 2000
  
```{r,echo=TRUE,warning=TRUE}
    library(tidyverse)
    library(ggplot2)
    library(dplyr)
    gapminder2000<-filter(gapminder, Year >= 2000)
``` 
1. How does income relate to life expectancy?

```{r,echo=TRUE,warning=TRUE}
    p1d<-aggregate(income ~ life, gapminder2000, median)
    library(ggplot2)
    p1<-qplot(life, income, data = p1d)
    p1
```
  
2. What is the population trend for United States vs India?
  
Filter and aggregate data for USA and India
  
```{r,echo=TRUE,warning=TRUE}
    gapminder2000AI <- filter(gapminder2000, Country == 'United States' | Country == 'India')
  
    p2d<-aggregate(Country ~ population, gapminder2000AI, median)
```

**Scatter plot of Population by Country**
```{r,echo=TRUE,warning=TRUE}
    library(ggplot2)
    p2<-qplot(Country, population, data = p2d)
    p2
```

**Box plot of Population by Country**
```{r,echo=TRUE,warning=TRUE}
    library(ggplot2)
    ggplot(p2d, aes(x = Country, y = population)) +
    geom_boxplot(outlier.colour = "hotpink") +
    geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4)
```

3. Top 10 and bottom 10 countries by income

Aggregate data

```{r,echo=TRUE,warning=TRUE}
    p3d<-aggregate(income ~ Country, gapminder2000, median)
```

**Top 10 countries by income**
  
```{r,echo=TRUE,warning=TRUE}
      top10a<-filter(p3d, row_number(desc(income)) <= 10)
    	top10<-arrange(top10a,income)
    	top10
```

**Bottom 10 countries by income**

```{r,echo=TRUE,warning=TRUE}
    bot10a<-filter(p3d, row_number(desc(-income)) <= 10)
    bot10<-arrange(bot10a,income)
    bot10
```

**Box plot of Top 10 highest income countries**

```{r,echo=TRUE,warning=TRUE}
    library(ggplot2)
    ggplot(aes(x = top10$Country, y = top10$income), data = top10) +
    geom_jitter(color = 'orange', alpha = 0.5, size = 0.5) +
    geom_boxplot(outlier.color='red', alpha = 0.5) +
    stat_summary(fun.y = "mean", geom = "point", color = "blue", shape     = 8, size = 4) +
    coord_cartesian() +
    ggtitle("Top 10 lowest income countries")
```

**Box plot of Top 10 highest income countries**

```{r,echo=TRUE,warning=TRUE}
   library(ggplot2)
    ggplot(aes(x = bot10$Country, y = bot10$income), data = bot10) +
    geom_jitter(color = 'orange', alpha = 0.5, size = 0.5) +
    geom_boxplot(outlier.color='red', alpha = 0.5) +
    stat_summary(fun.y = "mean", geom = "point", color = "red", shape     = 8, size = 4) +
    coord_cartesian() +
    ggtitle("Top 10 lowest income countries")
```

## Results and conclusion
  
#### Question 1: How does income relate to life expectancy?
  *There are more people in the 55 to 65 age range and the income falls below 3000.
  This correlates with the fact that the average life expectancy has been increasing over time so people live longer.
  There is a significant number of data pointsin the upper spectrum indicating richer people live longer as they can afford better medical facilities*
  
##### Question 1.1: How does income relate to life expectancy by region?
  *Countries from Sub Saharan Africa form the low end of the spectrum.
  Countries from Europe and America form the high end of the spectrum*
  
#### Question 2: What is the population trend for United States vs India?
  *The population is India has been high since the 1800s and has increased at a rapid pace.
  The population in United States is in the lower end and it's uptrend has been very slow and has a low people per square kilometer metric.*
  
#### Question 3: Top 10 and bottom 10 countries by income
  *Most countries in the top 10 countries with the highest GDP are from Europe.
  Oil rich countries like Qatar and Kuwait make it to the top 10 along with the Unites States
  Singapore is the only country from East Asia.
  All the countries in the bottom 10 are from Sub Saharan Africa.*
  
  